---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
---

Here is some analysis of steps daily activity of 'Activity monitoring data' dataset.

This work is done in order to fulfill Peer Assessment 1 of Reproducible Research Coursera's course.


## Loading and preprocessing the data


Load the csv data from zip file to data.table.
```{r load data, echo=TRUE}
library(data.table)

file    <- unz(description='activity.zip', filename='activity.csv')
rawdata <- read.csv(file=file)

rm(file)
```

Get rid of **NA** values; precalc step means over all days for each interval; precalc steps over each day.
```{r preprocessing, echo=TRUE}
tidydata <- data.table(rawdata[!is.na(rawdata$steps),])
tidydata.meansteps   <- tidydata[,list(mean=mean(steps)),by=interval]$mean
tidydata.stepsbydays <- tidydata[,list(sum=sum(steps)),by=date]$sum
```


## What is mean total number of steps taken per day?


TASK: *Make a histogram of the total number of steps taken each day.*
```{r hist of total steps, echo=TRUE, fig.width=12}
hist(tidydata.stepsbydays,
     breaks=20,
     axes=FALSE,
     xlab='Steps, #',
     ylab='Frequency, %',
     main='Total steps per day')
axis(side=1, 1000*rep(0:22), labels=FALSE, lwd=0.2)
axis(side=1, 5000*rep(0:5))
axis(side=2, 1:10, labels=FALSE, lwd=0.2)
axis(side=2, 2*0:5)
```

TASK: *Calculate and report the __mean__ and __median__ total number of steps taken per day.*
```{r mean and median, echo=TRUE}
median(tidydata.stepsbydays)
mean(tidydata.stepsbydays)
```


## What is the average daily activity pattern?


TASK: *Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).*
```{r daily average plot, echo=TRUE, fig.width=12}
plot(unique(tidydata$interval),
     tidydata.meansteps,
     type="l",
     axes=FALSE,
     xlab="5 min intervals",
     ylab="Average steps, #",
     main="Average daily activity pattern")
axis(side=1, 100*0:24, labels=FALSE, lwd=0.2)
axis(side=1, 500*0:5)
axis(side=2, 10*1:21, labels=FALSE, lwd=0.2)
axis(side=2, 50*0:5)
```

TASK: *Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?.*
```{r maxsteps interval, echo=TRUE}
tidydata[which.max(tidydata.meansteps)]$interval
```


## Imputing missing values


TASK: *Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).*
```{r missings stats, echo=TRUE}
mistat <- table(is.na(rawdata$steps))
names(mistat) <- c("has values", "missed values")
c(mistat[2],mistat[1])
```

TASK: *Create a new dataset that is equal to the original dataset but with the missing data filled in.*

All mised values will be filled with mean values.
```{r missings fulfilin, echo=TRUE}
cdata <- data.frame(tidydata[,list(mean=mean(steps)),by=interval])
cdata <- merge(cdata,rawdata[is.na(rawdata$steps),],)
cdata$steps <- cdata$mean
cdata <- cdata[c("steps","date","interval")]
cdata <- rbind(tidydata,cdata)
```

TASK: *Make a histogram of the total number of steps taken each day.*
```{r hist of corrected, echo=TRUE, fig.width=12}
hist(cdata[,list(sum=sum(steps)),by=date]$sum,
     breaks=20,
     axes=FALSE,
     xlab='Steps, #',
     ylab='Frequency, %',
     main="Total steps per day, NA's corrected")
axis(side=1, 1000*rep(0:22), labels=FALSE, lwd=0.2)
axis(side=1, 5000*rep(0:5))
axis(side=2, 1:20, labels=FALSE, lwd=0.2)
axis(side=2, 2*0:10)
```

TASK: *Calculate and report the mean and median total number of steps taken per day.*
```{r mean and median corrected, echo=TRUE}
cdata.stepsbydays <- cdata[,list(sum=sum(steps)),by=date]$sum
median(cdata.stepsbydays)
mean(cdata.stepsbydays)
```

TASK: *Do these values differ from the estimates from the first part of the assignment?*
```{r meanmedian comparison, echo=TRUE}
median(cdata.stepsbydays) - median(tidydata.stepsbydays)
mean(cdata.stepsbydays) - mean(tidydata.stepsbydays)
```

TASK: *What is the impact of imputing missing data on the estimates of the total daily number of steps?*

Imputing missing values based on median of existing values leads to increasing frequency of average value in dataset.

Mean and median are identical now.


## Are there differences in activity patterns between weekdays and weekends?


TASK: *Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.*
```{r weekday factor, echo=TRUE}
cdata.f <- weekdays(strptime(cdata$date,"%Y-%m-%d"))
library(plyr)
cdata.f <- as.factor(revalue(cdata.f, c("Sunday"="weekend",
                                        "Monday"="weekday",
                                        "Tuesday"="weekday",
                                        "Wednesday"="weekday",
                                        "Thursday"="weekday",
                                        "Friday"="weekday",
                                        "Saturday"="weekend")))
```

TASK: *Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).*
```{r weekday activity diff, echo=TRUE, fig.width=12}
library(lattice)
wdata <- cdata[,list(mean=mean(steps)),by=list(cdata.f,interval)]
par(mfrow=c(2,1))
with(wdata, {
    par(mar=c(0,5,1.5,0))
    plot(wdata[cdata.f=="weekday"]$interval,
         wdata[cdata.f=="weekday"]$mean,
         main="Weekday/end activity chart",
         axes=FALSE,
         type="l",
         ylab="Weekday steps, #")
    grid()
    axis(side=2, 10*1:24, labels=FALSE, lwd=0.2)
    axis(side=2, 50*0:5)
    par(mar=c(4.5,5,0,0))
    plot(wdata[cdata.f=="weekend"]$interval,
         wdata[cdata.f=="weekend"]$mean,
         axes=FALSE,
         type="l",
         xlab="5 min intervals",
         ylab="Weekend steps, #")
    grid()
    axis(side=2, 10*1:21, labels=FALSE, lwd=0.2)
    axis(side=2, 50*0:5)
    axis(side=1, 100*0:24, labels=FALSE, lwd=0.2)
    axis(side=1, 500*0:5)
},
)
```


## R Session Information


```{r session information, echo=TRUE}
sessionInfo()
```
